Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
more

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these generally cannot cope with universal quantifiers ranging over
infinite types. This paper introduces a translation that eliminates infinite
quantifiers in axioms specifying a large class of recursive functions,
including terminating, tail-recursive, and productive functions. An empirical
evaluation confirms that the approach works well on benchmarks from two
independent sources. The translation is implemented as a preprocessor in the
solver CVC4.
\end{abstract}

@@

Unfortunately, neither of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers. A simple
example where both approaches fail is
$$\forall n : {\sortint}.\;\, \lite\bigl(
n \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$$
This state of affairs is highly unsatisfactory, given the frequence of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-fun-rec}, to the SMT-LIB standard \cite{smtlib25}.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be
interpreted as $\sortint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0, \con{s}(x)
\teq 0, \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargsort{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$,
which means there exists a $\Sigma$-interpretation satisfying formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

>>>>

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these generally cannot cope with universal quantifiers ranging over
infinite types, such as integers and recursive datatypes.
This paper introduces a translation that eliminates infinite
quantifiers in axioms specifying a large class of recursive functions,
including terminating, tail-recursive, and productive functions. An empirical
evaluation confirms that the approach works well on benchmarks from two
independent sources. The translation is implemented as a preprocessor in the
solver CVC4.
\end{abstract}

@@

Unfortunately, neither of these fragments can accommodate the vast majority of
quantified formulas that correspond to recursive function definitions: The
essentially uninterpreted fragment does not allow the argument of a
recursive function to be used deep inside a complex term on the right-hand side,
whereas the finite model finding techniques of Reynolds et al.\ are not
applicable for functions over infinite domains such as the integers or
recursive datatypes. A simple example where both approaches fail is
$$\forall n : {\sortint}.\;\, \lite\bigl(
n \leq 0,\allowbreak\; \con{p}( n ) \teq 1,\allowbreak\; \con{p}( n ) \teq 2 * \con{p}( n - 1 ) \bigr)$$
This state of affairs is highly unsatisfactory, given the frequence of
recursive definitions in practice and the impending addition of a dedicated
command, \texttt{define-fun-rec}, to the SMT-LIB standard \cite{smtlib25}.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargsort{f}$ can be
interpreted as $\sortint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\sortint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargsort{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$,
which means there exists a $\Sigma$-interpretation satisfying formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

We presented a translation scheme that makes more SMT problems amenable to
finite model findings in SMT. In conjunction with the finite model finding
techniques introduced by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, it makes it possible to
find models of quantified formulas over infinite types, such as arithmetic
types and recursive datatypes.

in definitional form with respect to admissible function definitions

Universal quantifiers ranging over infinite types
